Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-05-02 paper.Rnw
Integrated Mallar's comments on the intro, and fixed references

%draft mode
\usepackage[colorlinks]{hyperref}
\usepackage[colorinlistoftodos, textwidth=4cm, shadow]{todonotes}
\usepackage[displaymath, tightpage]{preview}
\setlength{\marginparwidth}{1.2in}

@@

The hippocampus is of particular interest to many researchers because it is
implicated in forms of brain dysfunction such as Alzheimer's
disease\citep{Sabuncu2011} and schizophrenia\citep{Narr2004,Karnik-Henry2012},
and has functional significance in cognitive processes such as learning and
memory\citep{DenHeijer2012,Scoville2000}.  For many research questions
involving magnetic resonance imaging (MRI) data accurate identification of the
hippocampus and its subregions is a necessary first step to better understand
the individual neuroanatomy of subjects.

Currently, the gold standard for neuroanatomical segmentation is manual
delineation by an expert human rater.  This is problematic for hippocampal
segmentation for several reasons.  First, manual segmentation takes a
significant investment of time and expertise \citep{Hammers2003} which may not
be readily available to researchers or clinicians.  Second, the amount of data
produced in neuroimaging experiments increasingly exceeds the capacity for
identification of specific neuroanatomical structures by an expert manual
rater.  Third, the true definition of hippocampal anatomy in MR images is
disputed \citep{Geuze2004}, as evidenced by efforts to create an unified
segmentation protocol \citep{Jack2011}.

Compounding each of these problems is the significant neuroanatomical
variability in the hippocampus throughout the course of aging, maturation, and
neuropsychiatric disorders
\citep{Mouiha2011b,Sabuncu2011,Giedd1998,Gogtay2006,Narr2002}.  The result is that
existing hippocampal atlases available to a researcher may not accurately
represent neuroanatomy of a specific population under study.  Additionally, in
the course of a research or clinical study, it may be necessary to make
adjustments to hippocampal definition as a means of hypothesis testing.  For
example, Poppenk \citep{Poppenk2011} found that subdividing the hippocampus
into anterior and posterior regions resulted in a predictive relationship
between volume difference of those regions and recollection memory performance.
Making such modifications to a set of MRI data segmentations requires additional
manual effort.

Automated segmentation techniques do not require human intervention but do
require {\it a priori} anatomical information to guide segmentations.  In this
paper we focus on methods that use manually segmented MRI atlases as anatomical
priors, as these methods achieve some of the best automated hippocampal
segmentation accuracies to-date. This technique was first developed using a
single atlas prior (known as single-atlas, or model-based,
segmentation)\citep{Haller1997,Csernansky1998}.  Volumetric image registration
is used to estimate a fit between the neuroanatomy of an atlas and target
images.  Labelling of the target image is achieved by applying the resulting
transformation to the atlas labels to bring them into the target image space
({\em label propagation}). This method is limited in accuracy by the
introduction of estimation errors in registration and partial volume effects in
label resampling, and errors introduced when the anatomy of the atlas is
unrepresentative of the target anatomy.

Multi-atlas segmentation techniques address these limitations by combining
segmentation information from a series of expertly segmented atlases
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2012,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used merge
these labels into a definitive segmentation for the target.  In addition, {\em
atlas selection} techniques are often used to exclude atlases from label fusion
that are dissimilar to a target image in order to reduce error from
unrepresentative anatomy \citep{Aljabar2009}. Cross-correlation or normalised
mutual information of image intensities are common measures of image similarity
used in atlas selection.

Multi-atlas methods have been very successfully applied to hippocampal
segmentation. Collins et al. found near-manual segmentation performance using an
atlas library of 80 T1-weighted atlas images from the ICBM152 dataset, the
\animal nonlinear registration algorithm, normalised mutual information as an
atlas selection similarity metric, and majority vote for label fusion
\citep{Collins2010}.  The Alzheimer's Disease Neuroimaging Initiative (ADNI) is
a commonly used benchmarking dataset of MR images of controls and patients with
MCI or Alzheimer's (see Methods for more information on the ADNI dataset).
Leung et al. tuned parameters for registration and label fusion to the
segmentation of ADNI1 1 year dataset of images with an atlas library of 55
images \citep{Leung2010}.  The MAPER whole brain segmentation algorithm
\citep{Heckemann2006,Heckemann2011}, using 30 atlases, on all ADNI1 baseline
images \citep{Heckemann2011}.  Lotjonen et al. use images from the ADNI1
baseline dataset and 30 atlases with a proprietary non-linear registration
method based on intensity differences, and post-processing step using a graph
cuts algorithm to optimise fused segmentations against a spatial intensity prior
\citep{Lotjonen2012}.  
\todo{No story here}

The LEAP algorithm is an elegant modification to the basic multi-atlas strategy
\citep{Wolz2010}. The atlas library is grown, beginning with a set of manually
labelled atlases, and successively incorporats unlabelled target images after
being labelling using multi-atlas techniques. The sequence in which target
images are labelled is chosen so that the similarity between the atlas images
and the target images is minimised at each step, effectively allowing for
deformations between very dissimilar images to be broken up into sequences of
smaller deformations.  With an atlas library of 30 MR images, LEAP was used to
segment the ADNI1 baseline dataset, achieving a mean Dice score of 0.85 with
manual segmentations.

%- TODO: "probablistic methods?"

While not purely nulti-atlas techniques, there are several important algorithms
for hippocampal segmentation that inform our approach. The popular FreeSurfer
application's whole brain segmentation algoritm uses a probabilistic atlas of
anatomical and tissue classes along with spatial constraints for class labels
encoded using a Markov random field model \citep{Fischl2002}. When segmenting
hippocampal subfields, FreeSurfer employs a Bayesian inference algorithm using a
probabilistic atlas of anatomical classes as a prior, and a likelihood model of
how those classes translate into MR image intensities, both trained on manual
segmentations of high resolution MR images \citep{Fischl2009}. Yushkevitch et
al. describe a semi-automated method for hippocampal subfield segmentation of
focal T2 images\citep{Yushkevitch2010}.  The unlabelled MR image must be
manually partitioned into 'head', 'body' and 'tail', and then multi-atlas
methods are used to segment the image.  Finally, an AdaBoost-based bias
correction classifier is trained on texture, spatial location, and intensities
of manual segmentations and is applied to fix mislabelled voxels.

%  Chupin's use of shape-based priors 
%   - Mouiha2011b - Hippocampal atrophy rates in AD: compare SNT vs FreeSurfer
%     on ADNI
%

Aside from the algorithmic choices used in multi-atlas segmentation, it is
natural to ask about how the features of the atlases themselves impact the
resulting segmentations. As noted, by choosing atlases ranked most similar to a
target image by voxel intensity profile, segmentation accuracy is improved,
suggesting that neuroanatomical similarity plays a strong role
\citep{Aljabar2009}.  Carmichael et al. explored this directly and found that
when using only one atlas the important factors leading to improved accuracy are
that the atlas have neuroanatomical features that match the target, and that the
atlas segmentation use the same protocol as the gold-standard
\citep{Carmichael2005}. Nestor et al.  found that hippocampal segmentation
protocols that include more dorsal white-matter and posterior anatomy tended to
produce higher overlap and better accuracy at distinguishing disease classes in
the ADNI1 1 year dataset \citep{Nestor2012}. These results suggest both atlas
library neuroanatomy and delineation protocol play a significant role in the
resulting segmentation.

Considered along with our earlier discussion on the difficulty of producing
manual segmentations of MR images and the need for adaptable segmentation
definitions in order to conduct research, this presents a real problem of labour
and expertise when using existing multi-atlas segmentation methods which rely on
relatively large atlas libraries (typically between 30 and 80 atlases). Indeed,
it may be especially prohibitive to use these methods in situations where
producing a single atlas is challenging (e.g. histology-based atlases, or
atlases from very high resolution images). In this paper we address the problem
of producing accurate segmentations using small numbers of manually segmented
atlases.

% MAGeT bootstrapping: tune the template library to your dataset, reduce the
% number of input atlases you need, with the upshot that with fewer atlases it's
% that much easier to change definitions in order to test hypothesis

Our algorith, called \mb ({\em M}ultiple {\em A}utomatically {\em
Ge}nerated {\em T}emplates), is an extension to the basic multi-atlas-based
segmentation schema\citep{Chakravarty2012}. Principally, we explore the
possibility of using a small atlas library to bootstrap a much larger
\emph{template library} composed of images taken from the target population.
The template library is then used to segment the targets in a similar fashion
to basic multi-atlas segmentation: by label propagation and label fusion. The
intuition driving this approach is that by generating a template library we
leverage the unique neuroanatomy of target population on hand to initialize the
segmentation process and improve accuracy over direct propagation from the atlas
library to unlabelled targets while also using fewer manually segmented
atlases.

The insight of generating a template library is not new.  Heckemann et al.
compared ``indirect'' segmentation -- taking a single atlas and propagating the
labels to intermediate targets before fusing them in a target image space -- to
multi-atlas segmentation and found that the indirect approach performed worse
\citep{Heckemann2006}. In this paper we continue the same line of investigation
but explore the performance when using multiple atlases as well as the effect of
different registration and fusion methods.

The LEAP algorithm \citep{Wolz2010}, described above, is another example of
indirect segmentation previously explored.  LEAP proceeds by iteratively
segmenting unlabelled images most similar to the atlas library images and then
incorporating the labelled images into the atlas library for future iterations.
The novelty explored in our current work is to demonstrate the viability of
achieving comparable segmentation accuracy using the basic multi-atlas schema
and using significantly fewer manually created atlases.

In previous work \citep{Chakravarty2012}, we applied \mb to segmentation
of the human striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. The contribution of the present work is to extend
our approach to the human hippocampus and perform a series of experiments to
rigorous validate the method. First, we conduct an extensive cross-validation of
\mb and basic multi-atlas segmentation on a subset of the ADNI1 dataset to
assess the accuracy of \mb under various parameter settings (number of
atlases and templates, registration and fusion methods). With the best
performing parameter configuration discovered above, we estimate \mb
intra-rater reliability by segmenting separately acquired T1 images of the
atlas subjects. For this experiment, we use the Winterburn atlases: digital
hippocampal subfield segmentations of five {\em in vivo} high-resolution (300$\mu$
isotropic) T1-weighted MR scans\citep{Winterburn2013}. To validate \mb in a
real world situation, we segment the entire ADNI1 Complete 1Yr dataset and compare
our segmentations to established automated and manual segmentations.
Additionally, to ensure \mb accuracy across disease categories, we also
compare \mb segmentations to manual segmentations of 139 first episode
schizophrenic patients.

@@

\subsection{The \mb Algorithm}
In this paper we will be making a distinction between an atlas and a template --
typically these terms are used roughly interchangeably. The term {\it atlas} is
taken to refer to two image volumes: an intensity image ({\it atlas image}) and
a corresponding manual segmentation image ({\it atlas labels}). {\it Template}
refers more generically to any image and corresponding labelling, manual or
computed, when it is {\it used as} as a model in the segmentation of another
image.  The terms {\it atlas library} and {\it template library} mean a set of
such images. Additionally, we will use the terms {\it target} to refer to an
intensity image for which we would like an segmentation.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                       Registration Methods                          %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Registration Methods}

Before registration, all images underwent preprocessing with the N3 algorithm
\citep{Sled1998} to minimize intensity nonuniformity. In our experiments we use
one of two non-linear image registration methods.

\subsubsection{Automatic Normalization and Image Matching and Anatomical
Labeling (\animal)}

\subsubsection{Automatic Normalization Tools (\ants)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                           Label Fusion                              %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Label fusion methods}
Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this paper we explore three fusion methods.

@@

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI1 database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset download was the "ADNI1:Complete 1Yr 1.5T"
standardized dataset available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008a}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.

@@

@@

latex(snt.tests, #file="", 
      size="scriptsize",
      caption="", row.names="",
      title="")
@

>>>>

%draft mode
\usepackage[colorinlistoftodos, textwidth=4cm, shadow]{todonotes}
\usepackage[displaymath, tightpage]{preview}
\setlength{\marginparwidth}{1.2in}

@@

The hippocampus is a neuroanatomical structure situated in the medial temporal
lobe of the brain, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}.  In addition to its known functional roles,
the hippocampus is of interest to neuroscientists because it is implicated in
several forms of brain disfunction such as Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}.  In
neuroimaging experiments, magnetic resonance images (MRI) are often used for the
identification of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the unique neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human. However
the rapid increase in the availability of MRI data and the time and expertise
required for manual segmentation is prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007,Weiner2013}.
Further, there is little agreement between researchers regarding how exactly the
hippocampus should be identified in MRI images \citep{Geuze2004} and this has
led to efforts to create an unified hippocampal segmentation protocol
\citep{Jack2011}.

Automated segmentation techniques that are reliable, objective, and reproducible
are a necessary alternative to manual segmentation.  In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods.  The resulting nonlinear
transformation is applied to the manual labels (ie. {\em label propagation}) to
apply them to the target image.  While this methodology has been used
successfully in several contexts
\citep{MallarMedIA200809,MallarHBM20082009,Collins1995,Haller1997}, it is limited
in accuracy by the introduction of errors due inaccuracies in the nonlinear
transformation itself, partial volume effects in label resampling, and
irreconcilable differences between the neuroanatomy represented within the atlas
and target images.

One methodology that can be used to mitigate these sources of errors involves
the use of multiple manually segmented atlases and probabilistic segmentation
techniques, such as those found in the FreeSurfer package \citep{Fischl2002},
FreeSurfer uses a probabilistic atlas of anatomical and tissue classes along
with spatial constraints for class labels encoded using a Markov random field
model to segment the entire brain.

More recently, many groups have been using multiple atlases to improve overall
segmentation accuracy (ie. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into a definitive segmentation for the target.  In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such a cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations.  For example Collins and Pruessner \citep{Collins2010}
demonstrated that only 14 atlases, selected based on highest similarity between
medial temporal lobe neuroanatomy as evaluated by normalized mutual information
\citep{Studholme1999} from a library of 80 atlases, were required to achieve
accurate segmentations of the hippocampus.  Several methods have been explored
for label fusion including the STAPLE algorithm \citep{Warfield2004} that
computes a probabilistic segmentation using an expectation maximization
framework from an set of competing segmentations; or others where a subset of
segmentations can be estimated using metrics such as the sum of squared
differences in the regions of interest to be segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library; ranging from atlas libraries
that require between 30 \citep{Heckemann2006} and 80 \citep{Collins2010}
manually segmented atlases.  This strategy has the main drawback of being
inflexible as it does not easily accommodate varying the definition of the
hippocampal anatomy (such as the commonly used heuristic of subdividing the
hippocampus in to head, body, and tail \citep{Poppenk2011,Pruessner2000}).
Furthermore, none of these methods have demonstrated sufficient flexibility to
accommodate atlases that are somehow exceptional such as those derived from
serial histological data \citep{Chakravarty06a,Yelnik2007} or high-resolution MRI
data that enables robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}.  Due to
the recent availability of the latter, there has been increased interest in the
use of probabilistic methods for the identification of the hippocampal subfields
on standard T1-weighted images.  Our group recently demonstrated that through
use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data \citep{Chakravarty2013} is possible.  The
novelty of this manuscript is the extension our multi-atlas methodology to the
hippocampus using more than a single input atlas, while simultaneously limiting
the number of possible inputs used during segmentation, and demonstrating that
accurate identification of the hippocampal subfields is indeed possible using
this methodology.

There are few methods that have attempted to perform multi-atlas segmentation
with a limited number of input atlases. The LEAP algorithm is an elegant
modification to the basic multi-atlas strategy \citep{Wolz2010} in which the
atlas library is grown, beginning with a set of manually labelled atlases, and
successively incorporating unlabelled target images after themselves being
labelling using multi-atlas techniques. The sequence in which target images are
labelled is chosen so that the similarity between the atlas images and the
target images is minimised at each step, effectively allowing for deformations
between very dissimilar images to be broken up into sequences of smaller
deformations.  Although Wolz et al. begin with an atlas library of 30 MR images,
this method could theoretically work using a much smaller atlas library.  In
their validation, LEAP was used to segment the whole hippocampus in the ADNI1
baseline dataset, achieving a mean Dice score of 0.85 with manual segmentations.

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampal subfields using standard T1-weighted data.
The first is included with the FreeSurfer package \citep{VanLeemput2009}.  This
work is limited as it omits the tail of the hippocampus and the segmentation
protocol has yet to be fully validated.  Nonetheless, they demonstrate that the
applicability of their work using data from 10 subjects.  In the second method,
Yushkevich and colleagues \citep{Yushkevich2009} acquired and labelled
hippocampal subfields on high-resolution MRI data from post-mortem medial
temporal lobe samples an demonstrate it's applicability in the segmentation of
other MRI-volumes using nonlinear registration guided using manually derived
hippocampus masks and specific landmarks. In their work they demonstrate
accurate parcellations of the subfields.

Here we address the issue of limiting the number of input atlases by tuning our
algorithm, for segmentation of the entire hippocampus, using a multi-fold
experiment performed on a subset Alzheimer's Disease Neuroimaging Initiative
(ADNI) 1 dataset.  The tuning includes \todo{complete}. Based on the parameters
we find in our experiment, we validate our algorithm using all of the data
available in the ADNI Complete 1Yr sample and compare our segmentations to the
other segmentations that are available through the ADNI informatics portal.  To
ensure that we have not over-fit our parameters to the aging or
neurodegenerative brain, we also apply our segmentations to a dataset of normal
controls and individuals suffering from first episode psychosis. Finally, we
perform a leave-one-out validation experiment to determine if the subfields can
be accurately identified using our multi-atlas framework.

@@

\subsection{The \mb Algorithm}
In this paper we will make a distinction between an atlas and a template image.
The term {\it atlas} refers to two image volumes: an intensity image ({\it atlas
image}) and a corresponding manual segmentation image ({\it atlas labels}). The
term {\it template} refers to the {\it use} an any image and corresponding label
image (whether the labels are manual or computed) as a model in the segmentation
of another image.  The terms {\it atlas library} and {\it template library} mean
a set of such images.  Additionally, we will use the terms {\it target} to refer
to an intensity image for which we would like an segmentation.

@@

@@

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI1 database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset download was the "ADNI1:Complete 1Yr 1.5T"
standardized dataset available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.

@@

Before registration, all images underwent preprocessing with the N3 algorithm
\citep{Sled1998} to minimize intensity nonuniformity. In this experiment we use
one of two non-linear image registration methods.

\subparagraph{Automatic Normalization and Image Matching and Anatomical
Labeling (\animal)}

\subparagraph{Automatic Normalization Tools (\ants)}

\subparagraph{Label fusion methods}
Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this experiment we explore three fusion methods.

@@

latex(snt.tests, file="", 
      size="scriptsize",
      caption="", row.names="",
      title="")
@
